---
title: "POLI787 Lab 1: Advanced Research Pipelines in `R`"
format: html
author: "Audrey Sachleben"
date: today
---

```{r setup, include=FALSE}
# set output options for rendered file
knitr::opts_chunk$set(
  eval = TRUE, # evaluate code in rendered output
  echo = TRUE, # show code in rendered output
  include = TRUE, # include output in rendered output
  message = FALSE, # hide messages in rendered output
  warning = FALSE, # hide warnings in rendered output
  error = FALSE, # hide errors in rendered output
  cache = FALSE, # do not cache results between renders
  tidy = FALSE, # do not reformat code
  collapse = TRUE, # collapse code and console output in rendered output
  comment = "#>>>" # prefix for console output lines in rendered output
)
```

## Introduction to 787 Lab
Welcome to POLI 787 Lab! 

<!-- LABS ARE MANDATORY -- SYLLABUS SAYS THEY ARE DUE AT 5PM ON FRIDAY (ask santiago about this) -->
<!-- - i know last year they were not, but now they are! grinch! -->

basically say what we are going over...grab bag of different topics, that will hopefully be super applicable to you and your research...especially as your getting started working on your own projects from start to finish. having these tools will also make you a desirable coauthor, and so will having good clean reproducible code, which we'll cover today. topics we'll cover that you'll likely find useful are (not limited to) MLE, advanced modeling strategies, text-as-data, networks etc. 

SAYMAN'S LANGUAGE NEED TO CHANGE
The lab portion of POLI784 is designed to provide you with hands-on experience implementing the concepts and methods you learn about in the lecture portion of the class. We'll be doing this using the `R` programming language. To get you up and running for the semester, this week's lab will focus on introducing you to basic concepts and tools that you will encounter throughout the semester.

While this week will predominantly be structured like a guide / tutorial, future labs will revolve mostly around brief, structured exercises for you to solve. These exercises are mainly designed to give you some practice before having to tackle more involved statistical programming assignments on your problem sets.

## Office Hours and Problem Sets
SAYMAN'S TEXT NEED TO CHANGE
As you know, you will have to solve six graded problem sets for this course. Santiago has told you most of what you need to know about them already, but we have decided to add one additional element to support you in your learning this semester -- an oral check-in after each p-set. While you'll be submitting your p-sets in groups, the check-in will be a one-on-one with me. I am not yet completely set on how to do this logistically, but you'll likely be signing up for ~10min office hour slots after each p-set. These check-ins will be rather informal and ungraded, and their purpose is two-fold. First, you'll get a chance to discuss your solutions for the p-set and any challenges you faced while working on it. Essentially, you'll be able to get immediate feedback from me on how to solve issues that you may have encountered. Second, it is an incentive to not free-ride on the assignments, as having these check-ins will allow us to hold each individual group member to account when it comes to explaining the solutions submitted.

When working on the p-sets you should collaborate closely with your group members (and other students in the class, if desired). This will hopefully provide you with a chance to get help from and provide help to others, which should make solving the p-sets easier. At the same time, make sure to make ample use of office hours whenever you feel stuck on the assignments.

With that being said, please take the time to review the information below before solving the (very brief) exercise at the end of the file. Once you have read and worked through this file, you should be equipped with the basic knowledge and tools to tackle the programming tasks you will be presented with throughout the rest of the semester. At the same time, computers and software -- despite being deterministic beasts -- can sometimes be uncooperative ... we've all been there! So, if you find yourself struggling with anything at all related to the labs, please do not hesitate to reach out to me ASAP. A 5-minute debugging session during office hours can save you literal *hours* of time. The same goes for discussing the substantive content of the lecture itself too, of course. Make use of my (and Santiago's) office hours regularly -- it will make this semester a lot more manageable.


## AI in Lab

NO AI in lab...what you do on your own time (in accordance withthe syllabus) is up to you...but this is a good space to struggle with code and problem solve. if you don't understand your code, LLMs are not helpful and it's pathetic. 

SAYMAN'S TEXT NEED TO CHANGE
It might seem absurd to do this when you could also just write a brief prompt and get an LLM to write hundreds of lines of code for you in one go, but especially at the level of programming proficiency that tends to be common among 1st-year grad students, that approach often ends in tragedy. Getting large-scale output from an LLM requires the skills to properly *validate* that output, and that's precisely where things tend to go wrong. By writing pseudo code and getting code completions a few lines at a time, you are necessarily actively involved in reviewing and adapting the code (human in the loop!), which will not only help you spot mistakes but also force you to actively plan out your problem solving approach and closely engage with the output. This will help you learn from the LLM and teach you how to guide it, rather than letting it rob you of an opportunity to learn by letting it guide you.

## Have a Brainstorming activity (to get back into the semester) about what their current a current project pipeline for a research project looks like, aka waht steps they take (specifically) -- and have them write down 5-8 steps/things...or give them a sample research project and have them talk through the research pipeline steps they would take

- where do you store your code, where's your data stored? how do you name your files? etc. 
- how do you ensure your code is replicable?
- how do you work with coauthors?

## Research Pipelines in R
 have an introduction about how this is best practice, you should do this when you are starting to work on your own research (or if you already have your own research, you should start doing this)

basically this lab talks through project set up (enrivonments + benefits), debugging code, and profiling (or optimizing code -- diagnosing and figuring out how to make it run more smoothly)

also give a general reminder on good principles (that we are not going to go over)
- COMMENT YOUR CODE CLEARLY
- write clean code

- something that is really important of keeping track of your data (number of observations)
- knowing your data and checking your data to make sure your code does what you want it to do...and doing this as you go along

<!-- (merging congressional record speeches to speakers -- created duplicates etc (donald payne, people switching party)) -->

## `R` Projects
Whenever you are undertaking a new research project, it is good practice to create an `RStudio` [project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) for that entire research project. Your `RStudio` project can help streamline your research pipeline, and make your work easily reproducible. 

Working within an .Rproj has three main advantages: 

- Self-contained workspace: all scripts, data, and outputs for a specific paper (or dissertation chapter) are saved in a single place.

- Automatic working directory: opening an `R` project automatically sets your working directory to the project's root folder. This helps eliminate the need for the machine-specific *setwd()* commands.

- Easier collaboration and replication: the root directory is standardized, so all of your references to other folders or files (such as loading in data) can use relative file paths, which will let your code run more seamlessly on your co-author's machine or on a replication server. 

# Quick Exercise: Creating Lab 1 Project
Briefly, let's set up a project for today's lab. There are two ways to create a project, either through the `RStudio` menu or through the console. 

Option 1: Point and Click (`RStudio` Menu)

- File > New Project...

- Select New Directory (if you wish to make a 787 lab 1 folder)

- Select Existing Directory (if you have a 787 lab 1 folder and you want to link it)

- Click Create Project

Option 2: Via `RStudio` Console
Alternatively, you can use the *usethis* package to create a project directly from your console:
```{r}
# install.packages("usethis")
# usethis::create_project("file_path/POLI787_lab1") 
```

<!--(Following the discussion we just had, everybody needs a 787 folder, a 787 lab folder, and a 787 lab 1 folder) -->

## Organizing Your Project Directory
While creating an .Rproj file is a helpful first step, I've also found that sticking to a consistent subfolder structure within your project is what really helps keep your research pipeline organized. 

The exact folder names you choose are ultimately a matter of personal preference, but keeping the same structure across all your research projects will save you countless hours (especially if you're returning to a paper months later to address reviewer comments!). 

I've found the following structure to be highly effective. Like I said, what your subfolders in a project look like is up to you, but I think it helps to pick a style and be consistent: 


- Code/: Contains all .R, .Rmd, and .py, scripts. I also like to name these numerically to indicate the order in which they run (e.g., 1. pulling_record, 2. cleaning_files, 3. parsing_speeches)

- Data/: Contains all data for the project. I also find it helpful to keep your raw data separate from your clean working data within this folder

- Literature/: self-explanatory

- Paper/: Contains manuscript drafts (I typically save with the date if I'm making large changes rather than overwriting)

- Figures/: Contains all of your saved plots, graphs, regression tables, etc. Basically anything you create in `R` that's going into your paper is saved here

Tip: You can generate this structure in your console right after creating your project:

```{r}
# create standard sub-directories 
# purrr::walk(c("Code", "Data", "Literature", "Paper", "Figures"), dir.create)
```


## `R` Environments
One of the main benefits of projects is when they are used in conjunction with `Posit`'s [virtual environments](https://posit.co/blog/renv-project-environments-for-r/) for `R` via the [renv](https://rstudio.github.io/renv/articles/renv.html) package. While `RStudio` projects help organize your files, they do not manage your packages, which is what the renv package can do for you.

<!-- Ask how many are using renv, I know Sayman talked about it but didn't require it -->
Using renv replaces your global `R` library with a project-specific library. This provides two big advantages for researchers:

- Version control for packages: Using renv, your project relies only on the exact package versions you installed for that project. Sometimes, you may need to use multiple versions of a package across different projects. It may also be the case that packages from different projects are not compatible with one another (e.g. MASS package masks the 'select' command from dplyr). 
<!-- One project uses MASS another uses DPLYR and you open your new script to work on the dplyr one then you try to select columns in dplyr and it doesn't work -->

- Streamlining collaboration: renv tracks your dependencies in a renv.lock file. By sharing this file (via Github or even Dropbox), your co-authors can automatically download and install the exact `R` environment you were using. 

## Exercise: Building renv Pipeline
Let's implement the renv workflow for this project. This is good practice to implement with your own projects as well. Make sure you are currently inside the POLI787_Lab1 project we just created, and run through the following steps in your console:

1. Initialize the environment. For each project, you need to turn on renv. This command creates your project library and generates the first renv.lock file. 
```{r renv-init, eval=FALSE}
# install package, if not already present
if(!require("renv")) install.packages("renv")

# initialize renv for current project
renv::init()
```

2. Install packages. Because renv has been initialized, these installed packages will be loaded into your project-specific environment, rather than your global `R` environment There are two options for installing packages: 

- As you normally would, via "install.packages":
```{r, eval=FALSE}
# install the tidyr package
install.packages("tidyr")
```

- Through renv's specific package installation:
```{r}
# install the dplyr package
renv::install("dplyr@1.1.0") # this can be helpful for installing specific versions of packages
```

3. Snapshot the environment. Once you've installed packages (and your code is working), you need to save the state of your environment to the renv.lock file. You should do this periodically as you add new packages to your project. 
```{r}
# save the current state of your project's library
renv::snapshot()
```

4. Restore the environment. This is what is particularly useful when collaborating. If you were to take a turn working in a shared script, you would rely on the renv.lock file to make sure that you and your co-author are working in the exact same `R` environment. You do this using the "restore" command:
```{r}
# installs the exact package versions recorded in the renv.lock file for the project
renv::restore()
```

5. Updating Packages
If you ever need to update a packagewithin your renv environment, you can use the update function. Once you've confirmed that your code still runs with the updated package, you should take another snapshot to record the change to the environment. 
```{r}
# update a specific package
renv::update("dplyr")

# save new version of renv.lock file (after confirming code runs)
renv::snapshot()
```


## Git

```{r}
1 + 1
```

## Debugging Code
 
 [debug](https://bookdown.org/rdpeng/rprogdatascience/debugging.html)
 
talk about the importance of debugging code and being able to debug code (rather than just throw it in an LLM). introduce the debugging commands, go through a debugging exercise with code/that covers a topic that you should be familiar with from 784 labs
intro on debugging code and it's importance...talk about not AI

good first step is always to inspect the error, because sometimes it's super simple. For example, what is the error below?

```{r}
# try to calculate the mean of several values
meann(c(1, 2, 3, 4, 5))
```

Other times, however, debugging code can be a much more complicated process, especially if you're not exactly sure where in your code the error is coming from. In those instances, there are several functions that can help you debug your own code. traceback() displays the sequence of function calls that led up to the most recent error. debug() allows you to walk through a function one statement at a time, so you can inspect your environment after each execution to find the error. debugonce() does the same thing but only for the next call to that function. Finally, browser() allows you to choose a point in your code where the code should pause so that you can inspect the current environment and see if there are errors.

## Debugging Code Exercise

Now that you've learned about the debugging functions, let's put them into practice. As a review of material from POLI 784, we will simulate data and fit an OLS regression model, while debugging code along the way.

Rather than simulating all of our variables inside a single function, we will build our data simulation as a series of smaller functions. The argument each function takes is the number of observations you want to simulate. 

```{r}
# set seed for reproducibility
set.seed(9692)


### create function to simulate gender
simulate_gender <- function(N) {
  
  # simulate gender
  X1 <- sample(
    x = c(
      rep(0, N*0.5), # male
      rep(1, N*0.5) # female
    ),
    size = N,
    replace = FALSE
  )
  
  # return simulated gender
  return(X1)
}


### create function to simulate religious affiliation
simulate_religion <- function(N) {
  
  # create probabilities for religious affiliation
  rel_probabilities <- c(
    0.45, # protestant
    0.20, # catholic
    0.15, # jewish
    0.10, # muslim
    0.10 # other
  )
  
  # simulate religious affiliation
  X2 <- sample(
    x = c(1, 2, 3, 4, 5),
    size = N,
    replace = TRUE,
    prob = rel_probability
  )
  
  # convert religious affiliation to factor
  X2 <- factor(
    X2,
    levels = c(1, 2, 3, 4, 5),
    labels = c(
      "Protestant",
      "Catholic",
      "Jewish",
      "Muslim",
      "Other"
    )
  )
  
  # return simulated religious affiliation
  return(X2)
}


### create function to simulate demographic variables
simulate_demographics <- function(N) {
  
  # simulate biological sex
  X1 <- simulate_gender(N)
  
  # simulate religious affiliation
  X2 <- simulate_religion(N)
  
  # combine demographic variables
  demographics <- data.frame(
    X1 = X1,
    X2 = X2
  )
  
  # return simulated demographic variables
  return(demographics)
}


### create function to simulate years of education
simulate_years_education <- function(N) {
  
  # simulate years of education
  X3 <- sample(
    x = c(
      rep(11, N*0.10), # did not complete high school
      rep(12, N*0.35), # high school
      rep(16, N*0.35), # college
      rep(18, N*0.20) # college plus graduate
    ),
    size = N,
    replace = FALSE
  )
  
  # return simulated years of education
  return(X3)
}


### create function to indicate more than high school education
create_education_indicator <- function(X3) {
  
  # create indicator for graduate education
  X4 <- ifelse(
    df$X3 <= 18,
    0,
    1
  )
  
  # convert more than high school to factor
  X4 <- factor(
    X4,
    levels = c(0, 1),
    labels = c(
      "No Graduate Education",
      "Graduate Education"
    )
  )
  
  # return education indicator
  return(X4)
}

# create function to simulate education variables
simulate_education <- function(N) {
  
  # simulate years of education
  X3 <- simulate_years_education(N)
  
  # create indicator for more than high school
  X4 <- create_education_indicator(X3)
  
  # combine education variables
  education <- data.frame(
    X3 = X3,
    X4 = X4
  )
  
  # return simulated education variables
  return(education)
}


### create function to simulate age
simulate_age <- function(N) {
  
  # initialize age vector
  X5 <- rep(NA, N)
  
  # simulate age one observation at a time
  for (i in 1:N) {
    
    # randomly determine age distribution
    age_distribution <- sample(
      x = c(1, 2),
      size = 1,
      replace = TRUE,
      prob = c(0.75, 0.25)
    )
    
    # simulate younger age distribution
    if (age_distribution == 1) {
      
      X5[i] <- runif(
        n = 1,
        min = 18,
        max = 70
      )
      
    }
    
    # simulate older age distribution
    if (age_distribution == 2) {
      
      age_value <- rnorm(
        n = 1,
        mean = 70,
        sd = 10
      )
      
      if (age_value < 70) {
        
        age_value <- 
          age_value + 
          (2*(70 - age_value))
        
      }
      
      X5[i] <- age_value
      
    }
    
  }
  
  # round age to nearest year
  X5 <- round(X5)
  
  # return simulated age
  return(X5)
}


### create function to simulate outcome variable
simulate_outcome <- function(df) {
  
  # simulate outcome variable
  Y <- round(
    35 + 
    5*df$X1 +
    3*(df$X2 == "Protestant") + 
    5*(df$X2 == "Catholic") + 
    6*(df$X2 == "Jewish") + 
    (-5)*(df$X2 == "Muslim") +
    0.75*(df$X3 - 12) +
    0.25*df$X5 + 
    rnorm(
      n = nrow(df),
      mean = 0,
      sd = 12.5
    )
  )
  
  # return simulated outcome
  return(Y)
}


# create function to simulate full dataset
simulate_data <- function(N) {
  
  # simulate demographic variables
  demographics <- simulate_demographics(N)
  
  # simulate education variables
  education <- simulate_education(N)
  
  # simulate age
  age <- simulate_age(N)
  
  # combine simulated variables
  df <- data.frame(
    X1 = demographics$X1,
    X2 = demographics$X2,
    X3 = education$X3,
    X4 = education$X4,
    X5 = age
  )
  
  # simulate outcome variable
  df$Y <- simulate_outcome(df)
  
  # return simulated data
  return(df)
}
```

Run the function to simulate 3000 observations. When you encounter an error, use the traceback() function to inspect the function calls and find what led to the error. Once you've discovered the error, fix it in the function, and, using words, explain the error below the code chunk.
```{r}
# set sample size
N <- 3000

# simulate data
df <- simulate_data(N)

# examine the sequence of function calls that produced the previous error
traceback()

# simulate data again after fixing error
df <- simulate_data(N)

```

*Response*

Now that you've corrected that error, you are ready to fit an OLS model for your simulated data. Include X1, X2, X4, and X5 as predictors of Y.  
```{r}
# fit OLS regression model on simulated data
model1 <- lm(
  Y ~ X1 + X2 + X4 + X5,
  data = df
)


```

Oh no! You encountered another error. Turn on the debug() function and proceed with debugging the code. Fix the error and re-run the model. After you've re-run the model, 1) identify the error using words and 2) interpret the model coefficients. 

```{r}
# turn debugging on for simulation function
debug(simulate_data)

# run simulation function again, use debug() to proceed through and correct the error
df <- simulate_data(N)


# after correcting the error, turn debugging off
undebug(simulate_data)


# simulate corrected data
df <- simulate_data(N)

# fit regression model
model1 <- lm(
  Y ~ X1 + X2  + X4 + X5,
  data = df
)

# print model summary
summary(model1)

```

*Response*

## Profiling Code

talk about the importance of debugging code and being able to debug code (rather than just throw it in an LLM). introduce the debugging commands, go through a debugging exercise with code/that covers a topic that you should be familiar with from 784 labs

- make an exercise where students have to profile/then optimize code

```{r}
1 + 1
```




