---
title: "Maximum Likelihood and Binomial Models"
format: html
author: "Audrey Sachleben" 
date: "August 28, 2026"
---
```{r setup, include=FALSE}
# set output options for rendered file
knitr::opts_chunk$set(
  eval = FALSE, # evaluate code in rendered output
  echo = TRUE, # show code in rendered output
  include = TRUE, # include output in rendered output
  message = FALSE, # hide messages in rendered output
  warning = FALSE, # hide warnings in rendered output
  error = FALSE, # hide errors in rendered output
  cache = FALSE, # do not cache results between renders
  tidy = FALSE, # do not reformat code
  collapse = TRUE, # collapse code and console output in rendered output
  comment = "#>>>" # prefix for console output lines in rendered output
)
```

## Binomial Models

This week in lab we are shifting towards modeling with binary outcomes, or outcomes that can take on only one of two mutually exclusive values. Binary outcomes are exremely common in political science, so knowing how to work effectively with them is critical. 

From POLI 784 you already know how to estimate a linear regression using lm(). So, one reasonable starting point might be to wonder why we wouldn't also use OLS when our outcome variable is binary. As Santiago discussed in lecture, you actually can, however this creates some important problems. We'll use those problems to motivate two common binomial models, the logit and the probit, and learn how to estimate them using `glm()`. 

Finally, we'll also look under the hood at how these models are estimated, and we'll use `optim()` to manually estimate the same logit model we fit using glm(). 

## Script Set Up
```{r set up}
# clear environment
rm(list = ls())

# load packages
library(haven)

# load lab 2data
colombia <- read_dta("lab2.dta")

# set seed
set.seed(9692)

```

## Starting with a Linear Probability Model

Suppose we have a small survey in which `interest` measures someone's level of political interest (0 to 6) and `turnout` equals 1 if the respondent voted and 0 if the respondent did not vote. We could estimate the relationship between political interest and turnout using the `lm()` command you are familiar with. When OLS is used with a binary dependent variable, we generally call it a *linear probability model (LPM)*.

*Note*: The data below are fake and are intentionally constructed to make a point, so don't read anything substantive into them!

```{r sim data}
# create fake turnout data
turnout_data <- data.frame(
  interest = rep(0:6, each = 10),
  turnout = c(
    rep(0, 10),
    rep(0, 10),
    c(rep(0, 9), 1),
    c(rep(0, 5), rep(1, 5)),
    c(0, rep(1, 9)),
    rep(1, 10),
    rep(1, 10)
  )
)

# examine turnout across levels of political interest
table(__________, ____________)

# estimate linear probability model
lpm <- _____________

summary(lpm)

```

As you can see, one appealing feature of the LPM is that its coefficients are straightforward to interpret. Because `turnout` is coded 0/1, the coefficient on `interest` gives us the estimated change in the probability of turnout associated with a one-unit increase in political interest.

**Question:** According to the model, how does a one-unit increase in political interest change the predicted probability of voting?

*Response:*

While this may seem quite convenient at first glance, Santiago discussed how one problem associated with a linear probability model is that it can produce predicted probabilities that are negative or exceed 1. Let's take a look:

```{r lpm}
# generate predicted values from linear probability model
turnout_data$lpm_prediction <- predict(__________)

# inspect summary of predicted values
________(_____________)

```

This problem is particularly pronounced at the extremes of our data (those with an interest of 0 or 6). As you can see, OLS does not know that our dependent variable represents a binary outcome. It simply fits the best linear relationship through the data, and nothing prevents that line from extending below 0 or above 1. This is one reason we often turn to models specifically designed for binary outcome variables. 

## Generalized Linear Models

One way to estimate models for binary outcomes in `R` is with the `glm()` command. The basic syntax for a glm() should look pretty familiar. 
```{r glm syntax, eval = FALSE}
glm(
  outcome ~ predictor,
  data = data,
  family = ...(link = "...")
)
```

The important new piece is the *family* argument. The family tells `R` about the distribution of out outcome variable, and the *link function* we want to use. For a binary dependent variable, we use the binomial family. Two of the most common link functions are *logit* and *probit*. Let's estimate a logit model using our fake turnout data. *Note*: When you inspect the output from the glm model, you'll notice values such as "null deviance" and "AIC." We'll talk more about those next week when we focus on model evaluation and interpretation. 

```{r turnout logit}
# estimate logit model on turnout data
turnout_logit <- ______(
  ________ ~ ________,
  data = ________,
  ________ = ________(________ = "________")
)

summary(turnout_logit)
```

Now generate predicted probabilities from the logit model. We can then compare those the the predictions gathered from the linear probability model we previously estimated. The argument `type = "response"` tells `predict()` that we want predictions on the scale of our dependent variable—in this case, probabilities. As you'll see, the logit model (and probit) produces predicted probabilities that fall between 0 and 1. 

```{r pred logit}
# generate predicted probabilities for logit model
turnout_data$logit_prediction <- ________(
  ________,
  type = "response"
)

# compare summary of predictions between lpm and logit models
________(________$________)
________(________$________)
```

# Binomial Models Exercise: Paramilitary Presence in Colombia
Now, let's work through a real example. For the rest of today's lab, we'll use data from a previous project examining paramilitary groups in Colombia (Chelsea Estancona '18). Our dependent variable is `AUC`, which measures the presence of the *Autodefensas Unidas de Colombia (AUC)*, or United Self-Defenses of Colombia, in a municipality.

Our predictors are:

- `tribut_2_log`: logged tax income
- `tribut_2_log_sq`: logged tax income squared
- `pop_attacks`: attacks on the population
- `disbogota`: distance from Bogotá
- `drugs`: presence of drug crops
- `pop_log`: logged population 

First, inspect the data to make sure the outcome variable(`AUC`) is actually binary. Then, estimate a logit model and a probit model, respectively. Inspect the coefficient estimates from the logit and the probit side by side. 

```{r colombia glm}
# inspect dependent variable
table(colombia$AUC)

# estimate logit model
model_logit <- ________________

# examine model results
summary(model_logit)

# estimate probit model
model_probit <- __________________

# examine model results
summary(model_probit)

# compare logit and probit coefficient estimates
cbind(
  Logit = coef(model_logit),
  Probit = coef(model_probit)
)

```

The coefficients from the logit and probit models are clearly not numerically identical. That is okay! Logit and probit use different link functions, so the coefficient estimates operate on different scales. The direction of the estimates and the general substantive patterns of the relationships between predictors and the outcome should remain relatively similar, even though the estimates are different. 

**Question:** Compare the logit and probit models. Do any predictors change direction across the two models? Would you reach dramatically different substantive conclusions from them?

*Response:*

## Interpreting Logit & Probit Models

Logit and probit models give us predicted probabilities that stay between 0 and 1. However, the coefficients are no longer as straightforward to interpret as the coefficients from out linear probability model. For a logit model, the coefficients represent changes in the *log-odds* of the outcome: 

$$
\log\left(\frac{p_i}{1-p_i}\right)=X_i\beta.
$$

Log-odds aren't exactly the most intuitive quantity to communicate, so one thing we can do is exponentiate the coefficients to convert them into *odds ratios*. Odds ratios have a useful reference point:

- An odds ratio of 1 indicates no change in the odds.
- An odds ratio greater than 1 indicates increasing odds.
- An odds ratio less than 1 indicates decreasing odds.

Why does exponentiating a logit coefficient give us an odds ratio? Recall that our logit model can be written as:

$$
\log\left(\frac{p}{1-p}\right)=\alpha+\beta X.
$$

If we exponentiate both sides, the predicted odds are:

$$
\frac{p}{1-p}=e^{\alpha+\beta X}.
$$

Now suppose we increase $X$ by one unit. The ratio of the new odds to the original odds is:

$$
\frac{e^{\alpha+\beta(X+1)}}{e^{\alpha+\beta X}}
=
e^\beta.
$$

Everything else cancels out! This is why exponentiating a logit coefficient gives us the odds ratio associated with a one-unit increase in that predictor.


Now let's apply this, focusing on the logit model we just fit. Inspect and interpret the estimate on `pop_attacks`. 

```{r odds ratio}
# inspect logit coefficient for attacks on population
coef(model_logit)["pop_attacks"]

# convert coefficient into odds ratio
________(coef(model_logit)["pop_attacks"])

# calculate percentage change in odds
100 * (
  exp(coef(model_logit)["pop_attacks"]) - 1
)
```


**Question:** Holding the other variables constant, how is an additional attack on the population associated with the odds of AUC presence?

*Response:*

It's important to remember that odds are **not** probabilities. Odds ratios can be useful, but we often ultimately want to say something about the predicted probability of observing our outcome. We'll spend time next week generating predicted probabilities using the *average case approach* and the *observed value approach*, but for now, we'll focus on using the `predict()` and `plogis()` functions. First, use `predict()` to generate the model's linear predictor. For the logit model, this is the predicted log-odds. To transform them into probabilities, we can use the `plogis()` function. 

```{r log odds}
# generate predictions on log-odds scale
logit_linear_prediction <- __________(
  model_logit,
  type = "link"
)

# inspect first few values
head(logit_linear_prediction)

# transform log-odds into probabilities
logit_probability <- __________(
  ____________
)

# inspect first few probabilities
head(logit_probability)

# inspect range
summary(logit_probability)

```


Notice that all of these values fall between 0 and 1. This is exactly what `R` was doing earlier when we specified *predict(model_logit, type = "response")*. So, `plogis()` is the *inverse-logit transformation*. It takes an unbounded linear prediction and maps it onto a valid probability between 0 and 1. The same basic process occurs with a *probit* model. The difference is that the probit uses the CDF of a standard normal distribution rather than the logistic distribution. To estimate predicted probabilities for a probit model, we therefore use `pnorm()` rather than `plogis()`.


## Maximum Likelihood Estimation

So far, `glm()` has done all of the estimation for us. We told `R` which dependent variable we wanted to model, which predictors we wanted to include, and which link function to use, and `glm()` returned a set of coefficient estimates.

But where did those particular coefficient values come from?

When estimating generalized linear models, we often have one or more unknown parameters that need to be estimated. More generally, suppose we observe a set of outcomes $y_i$ and want to estimate some unknown parameter, or set of parameters, $\theta$. The likelihood function tells us how likely we are to observe our data for a particular value of $\theta$:

$$
p(y \mid \theta) = L(\theta \mid y).
$$

If we have a set of observations $y_1, y_2, \ldots, y_n$, the likelihood of observing the entire dataset is the product of the likelihood contributions from each individual observation:

$$
L(\theta \mid y_1,y_2,\ldots,y_n)
=
\prod_{i=1}^{N} p(y_i \mid \theta).
$$

Multiplying a large number of probabilities together can become computationally difficult. So, instead, we generally work with the *log-likelihood*. As you know, taking the log turns the product into a sum:

$$
\log L(\theta)
=
\sum_{i=1}^{N}
\log p(y_i \mid \theta).
$$

Our task is then to find the value of $\theta$ that maximizes this log-likelihood:

$$
\hat{\theta}
=
\underset{\theta}{\operatorname{argmax}}
\sum_{i=1}^{N}
\log p(y_i \mid \theta).
$$

At the end of the day, *maximum likelihood estimation (MLE) is therefore an optimization problem*. We are looking for the set of parameter values that makes the data we actually observed as likely as possible.

For some models, we can solve for these estimates analytically. For many others, there is no convenient closed-form solution, so we need to find the maximum numerically. One way to do this is the *Newton-Raphson method*, which we discussed in lecture and uses information from the first and second derivatives of the likelihood function to iteratively search for the parameter values that maximize it.

Fortunately, we do not have to program Newton-Raphson ourselves today! `Optim()` can do that for us!

Returning to our example of paramilitary groups in Colombia, for our logit model each observation has some predicted probability $p_i$ of having `AUC = 1`. The likelihood contribution for observation $i$ is:

$$
p_i^{y_i}(1-p_i)^{1-y_i}.
$$

If $y_i = 1$, this reduces to $p_i$. If $y_i = 0$, it reduces to $1-p_i$. The log-likelihood for our binary outcome can therefore be written as:

$$
\log L(\beta)
=
\sum_{i=1}^{N}
\left[
y_i\log(p_i)
+
(1-y_i)\log(1-p_i)
\right].
$$

Our goal is to find the values of $\beta$ that maximize this quantity.

Usually, convenient functions like `glm()` do all of this work for us. But we can also optimize a likelihood function ourselves using `optim()`.

## Optimization

The `optim()` function is a general-purpose optimization function in `R`. Unlike `glm()`, `optim()` has absolutely no idea that we specifically want to estimate a logit model. We're going to have to tell it exactly what function we want it to optimize. For this example, our goal is to reproduce the same logit model we already estimated using `glm()`, but this time we'll do it using `optim()`. 

First, we'll extract the outcome and design matrix that `glm()` actually used. 

```{r optim1}
# extract data used to estimate logit model
model_data <- model.frame(model_logit)

# extract dependent variable
y <- model.response(model_data)

# create design matrix
X <- model.matrix(model_logit)

# inspect first few rows
head(X)
```

Notice that the design matrix includes a column of 1s for our intercept along with each of our predictor variables. We next need to write the function we want `optim()` to optimize. There are a few piece worth thinking about here. First, `eta <- X %*% beta` calculates $X\beta$ then ``p <- plogis(eta)` converts those predicted log-odds into probabilities. Finally, the `dbinom()` function calculates each observation's contribution to the binomial log-likelihood. The syntax for optim is below:

*Note*: optim()` minimizes a function by default. We want to maximize the log-likelihood, so instead we minimize the negative log-likelihood. The parameter values that accomplish those two things are the same. 

```{r optim2}
# write negative log-likelihood function
logit_nll <- function(beta, y, X) {

  # calculate linear predictor
  eta <- X %*% beta

  # transform linear predictor into probabilities - what function does this?
  p <- plogis(eta)

  # calculate negative log-likelihood
  nll <- -sum(
    dbinom(
      y,
      size = 1,
      prob = p,
      log = TRUE
    )
  )

  # return negative log-likelihood
  return(nll)
}
```

# `Optim()` Exercise

Now, let's optimize! `optim()` has several important arguments:

- `par`: our starting values for the parameters
- `fn`: the function we want to optimize
- `method`: the optimization algorithm
- `hessian`: whether we want `R` to return the Hessian matrix

We'll use `"BFGS"`, which is a quasi-Newton optimization method. We'll use `optim()` to try to estimate the same glm model that we fit in the previous example about paramilitary groups in Colombia. 

```{r optim3}
# create starting values
starting_values <- rep(
  0,
  ncol(X)
)

# estimate logit model using optim()
optim_logit <- optim(
  par = ________,
  fn = __________,
  y = y,
  X = X,
  method = "BFGS",
  __________ = ________
)

# inspect optimization results
optim_logit

# extract coefficient estimates
____________

# compare glm and optim coefficient estimates
cbind(
  GLM = __________,
  OPTIM = __________
)
```


The two sets of estimates should be extremely close. That is the main point of this exercise! *Note*: the reason why the values are not identical is because of the starting values we selected when fitting `optim()`. 

We didn't estimate some new model using `optim()`. We manually told `optim()` to estimate the *same logit model* that `glm()` had already estimated for us.

# Standard Errors 

In the above code, we set `hessian = TRUE` when we called `optim()`. The Hessian contains information about the curvature of our likelihood function around its optimum. We can use its inverse to recover our estimated variance-covariance matrix and, from there, our standard errors. We can then compare the standard errors from our previoulsy estimated logit model and the optim()function we just ran.

```{r optim4}
# calculate standard errors from optim hessian
optim_se <- sqrt(
  diag(
    solve(optim_logit$hessian)
  )
)

# compare standard errors
cbind(
  GLM = sqrt(diag(vcov(model_logit))),
  OPTIM = optim_se
)
```

Again, the values should be very similar.

# Flexibility of `optim()`

Given this exercise, it might be reasonable to think, "Why did we just write all of that code when `glm()` estimated the same model in about three lines?" For a standard logit or probit model, you probably *should* use `glm()`. 

The main advantage of `optim()` is its *flexibility*. `glm()` knows how to estimate a particular set of generalized linear models. If the model you want fits within that framework, great. But suppose we wanted to specify a model in which not only the expected value of $Y$ depends on predictors, but the *variance itself also has a regression model*. For example:

$$
Y_i \sim N(\mu_i,\sigma_i^2)
$$
where:
$$
\mu_i = X_i\beta
$$
and:
$$
\log(\sigma_i^2)=Z_i\gamma.
$$

Now we have one set of coefficients predicting the mean and another set of coefficients predicting the variance. A standard call to `glm()` does not allow us to specify an arbitrary second regression for the observation-specific variance. But if we can write down the likelihood for that model, we can write a function containing both $\beta$ and $\gamma$ and pass it to `optim()`.

This is the kind of application where knowing how to use a general optimizer becomes particularly useful. You won't normally want to reinvent a model for which a reliable `R` package already exists. But understanding `optim()` gives you the ability to estimate models that don't fit neatly into the convenient functions you're used to. 
