---
title: "Lab 3: Model Evaluation and Interpretation"
format: html
author: "Audrey Sachleben" 
date: "September 4, 2026" 
---

```{r setup, include=FALSE}
# set output options for rendered file
knitr::opts_chunk$set(
  eval = FALSE, # evaluate code in rendered output
  echo = TRUE, # show code in rendered output
  include = TRUE, # include output in rendered output
  message = FALSE, # hide messages in rendered output
  warning = FALSE, # hide warnings in rendered output
  error = FALSE, # hide errors in rendered output
  cache = FALSE, # do not cache results between renders
  tidy = FALSE, # do not reformat code
  collapse = TRUE, # collapse code and console output in rendered output
  comment = "#>>>" # prefix for console output lines in rendered output
)
```

## Model Evaluation and Interpretation

Last week, we focused on estimating models for binary dependent variables. We used `glm()` to estimate logit and probit models and discussed why the raw coefficients from these models are not directly interpretable as changes in probability.

Today, we're going to shift from estimating a model to thinking about what we actually do with that model once we have it. We'll focus on two related tasks.

First, we'll work on model interpretation. We'll use simulation to calculate predicted probabilities and incorporate uncertainty in our coefficient estimates. In doing so, we'll compare two approaches to generating predictions: the *average-case approach* and the *observed-case approach*.

Second, we'll turn to model evaluation. We'll use the Receiver Operating Characteristic (ROC) curve and the area under that curve (AUROC) to evaluate how well our model distinguishes between observations for which the outcome occurred and observations for which it did not.

Throughout the lab, we'll try to explain the following question: what predicts whether congressional candidates discuss immigration on their campaign websites?

```{r set up}
# clear environment
rm(list = ls())

# load packages -- install if necessary
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
#install.packages("MASS")
library(MASS)
#install.packages("AUC")
library(AUC)

# set seed
set.seed(9692)
```

We'll work with data on U.S. House candidates and their campaign websites. Each observation is a candidate running for Congress between 2018 and 2022. Our dependent variable, `immigration`, equals 1 when a candidate discusses immigration somewhere on their campaign website and 0 when they do not. Read in the data and select the relevant variables for today's lab. 

```{r data}
# load lab 3 data
lab_data <- read.csv("lab3_data.csv")

# retain variables needed for today's lab
lab_data <- lab_data %>%
  dplyr::select(
    immigration,
    quality_cand,
    contested_primary,
    district_lean,
    log_fundraising,
    cand_party,
    male
  ) %>%
  drop_na()

# convert to factor variables
lab_data <- lab_data %>%
  mutate(
    quality_cand = factor(
      quality_cand,
      levels = c(
        "Inexperienced Challenger",
        "Experienced Challenger",
        "Incumbent")
      ),
    cand_party = factor(cand_party),
    male = factor(male)
  )

```

Next, fit a logistic regression predicting whether a candidate discusses immigration. Our primary predictor of substantive interest will be `district_lean`. This variable measures the share of the presidential vote received by the candidate's party in the district. Higher values therefore indicate that a candidate is running in a district that is more favorable to their own party.

The other predictors are:

- `quality_cand`: whether the candidate is an inexperienced challenger, experienced challenger, or incumbent

- `contested_primary`: whether the candidate faced an opponent in the primary

- `log_fundraising`: logged pre-primary fundraising

- `cand_party`: Democratic or Republican candidate

- `male`: candidate gender

```{r, eval = FALSE}

# fit logit model
model_logit <- ______(
  ______ ~ 
    ______ 
    + ______ 
    + ______ 
    + ________ 
    + ______ 
    + ________,
  data = lab_data,
  ______ = ________(______ = "______")
)

# inspect the output
summary(model_logit)

```

 Today, we're going to spend much less time looking at the coefficient table that is produced using `glm()`. Instead, we'll treat the fitted model as our starting point and ask what quantities we can calculate from it.

Look specifically at the coefficient on `district_lean`.

**Question:** What can we learn immediately from this coefficient? What can we *not* learn immediately from the coefficient?

*Response:* 

Recall that a logit coefficient represents a change in log-odds, not a constant change in probability. We can use the coefficient to determine the direction of the relationship, but it is often easier to communicate the substantive relationship using predicted probabilities.

## Coefficients to Predicted Probabilities

Before we introduce simulation, let's first generate a predicted probability for every candidate in our estimation sample.

```{r, eval = FALSE}
# generate fitted probabilities for candidates in estimation sample
predicted_probabilities <- ________(
  ________,
  ______ = "________"
)

# inspect range of predictions
summary(predicted_probabilities)
```

Each value is the model's estimated probability that the corresponding candidate discusses immigration, given that candidate's observed characteristics.

These predictions are based on the single set of coefficient estimates returned by `glm()`. But those coefficients are themselves estimates and therefore contain uncertainty. If we want to communicate uncertainty around our predicted probabilities, one useful strategy is simulation.

## Simulating Predicted Probabilities

Our fitted model gives us a vector of coefficient estimates, $\hat{\beta}$, and a variance-covariance matrix describing the uncertainty in those estimates.

Rather than treating the estimated coefficients as known with certainty, we can draw many sets of coefficients from their approximate sampling distribution. For each simulated set of coefficients, we then calculate the predicted probability we care about.

First, set a seed for reproducibility and choose the number of simulations. We'll use 1,000. 

```{r, eval = FALSE}
# set seed for reproducibility
set.seed(9692)

# number of simulations
n_sim <- ______
```

We'll use `mvrnorm()` from the `MASS` package to draw 1,000 simulated coefficient vectors from a multivariate normal distribution centered on our estimated coefficients. The function takes three main arguments:

- `n` specifies the number of simulated coefficient vectors we want to draw

- `mu` specifies the mean of the multivariate normal distribution. We set this equal to our estimated coefficients from the logit model.

- `Sigma` specifies the variance-covariance matrix of the multivariate normal distribution. We use `vcov(model_logit)`, which contains both the estimated variance of each coefficient and the covariance between coefficients.

Using the full variance-covariance matrix is important because the uncertainty in our coefficient estimates is not necessarily independent. `mvrnorm()` preserves these relationships when drawing sets of coefficients.

```{r, eval = FALSE}
# simulate coefficient vectors
beta_sims <- MASS::______(
  n = ______,
  mu = ______(________),
  Sigma = ________(______)
)

# inspect dimensions
dim(beta_sims)

# inspect the mean coefficients for the simulated data
colMeans(beta_sims)

```

Each row of `beta_sims` is one simulated set of coefficients, while each column corresponds to one coefficient in our fitted model.

Because we drew these simulated coefficients from a multivariate normal distribution centered on the coefficient estimates from `model_logit`, the average value of each simulated coefficient should be very close to the corresponding coefficient from the original model. However, the simulated means will generally not be exactly identical to the original estimates because we are drawing a finite number of random samples from that distribution. If we increased the number of simulations substantially, the simulated means would generally move even closer to the estimated coefficients.

**Question:** Why do we simulate the coefficients rather than calculate predicted probabilities using only the coefficients from the logit mdoel (`coef(model_logit)`)?

*Response:* 

The key idea is that we can now calculate our quantity of interest under 1,000 simulated versions of our fitted model. The distribution of those simulated quantities gives us a way to summarize both our point estimate and our uncertainty. 

We'll use these same coefficient simulations with two different approaches to generating predicted probabilities.

## Average-Value Approach

One traditional way to interpret a nonlinear model is to construct a hypothetical observation. The basic idea is to:

1. Pick the predictor we care about.
2. Allow it to vary over a meaningful range.
3. Hold the other variables at selected values.
4. Calculate the predicted probability at each value.

Our predictor of interest is `district_lean`. Let's create a sequence covering its observed range in the estimation sample.

```{r}
# create sequence over observed range of district lean
district_vec <- seq(
  from = min(lab_data$district_lean),
  to = max(lab_data$district_lean),
  length.out = 100
)
```

Now we need to decide who our hypothetical candidate is.

Importantly, the *average-value approach does not mean that we should blindly take the mathematical mean of every variable.* What would it mean to have a candidate who is 47% Republican or 0.31 incumbents? For categorical variables, we have to select an actual category.

For this example, we'll examine an inexperienced Democratic male challenger facing a contested primary. We'll hold fundraising at its sample mean and vary only `district_lean`.

```{r}
# construct hypothetical candidates across values of district lean
hyp_data <- data.frame(
  quality_cand = factor(
    rep("Inexperienced Challenger", length(district_vec)),
    levels = levels(lab_data$quality_cand)
  ),
  contested_primary = 1,
  district_lean = district_vec,
  log_fundraising = mean(lab_data$log_fundraising),
  cand_party = factor(
    rep("Democrat", length(district_vec)),
    levels = levels(lab_data$cand_party)
  ),
  male = factor(
    rep("Male", length(district_vec)),
    levels = levels(lab_data$male)
  )
)

# inspect first few hypothetical observations
head(hyp_data)
```

Next, convert these hypothetical candidates into the same design-matrix structure that `glm()` used to estimate the model.

```{r}
# create design matrix for hypothetical candidates
hyp_matrix <- model.matrix(
  ~ quality_cand + contested_primary + district_lean +
    log_fundraising + cand_party + male,
  data = hyp_data
)

# make sure columns match fitted model coefficients
hyp_matrix <- hyp_matrix[
  ,
  names(coef(model_logit)),
  drop = FALSE
]

# inspect dimensions
dim(hyp_matrix)
```

**Question:** Why do we need to convert `hyp_data` into a model matrix before multiplying it by our simulated coefficients?

*Response:* 

Now we can calculate the predicted probability for each hypothetical candidate under each simulated set of coefficients.

Remember that multiplying the design matrix by our coefficients gives us the linear predictor, $X\beta$, which is expressed in log-odds for a logit model. We use `plogis()` to transform those values back into probabilities.

```{r, eval = FALSE}
# calculate simulated predicted probabilities
average_pp <- ________(
  hyp_matrix %*% t(beta_sims)
)

# inspect dimensions
dim(average_pp)
```

The rows of `average_pp` correspond to values of `district_lean`, while the columns correspond to our 1,000 simulated coefficient vectors. We can now summarize those simulations at each value of `district_lean`.

```{r, eval = FALSE}
# summarize simulated predicted probabilities
average_output <- data.frame(
  district_lean = district_vec,
  # average predicted probability across simulations
  pp_avg = apply(
    ______,
    1,
    mean
  ),
 # lower bound of 95% confidence interval
  lb = apply(
    ______,
    1,
    quantile,
    probs = 0.025
  ),
  # upper bound of 95% confidence interval
  ub = apply(
    ________,
    1,
    quantile,
    probs = 0.975
  )
)

# inspect output
head(average_output)
```

Finally, plot the predicted probabilities and simulation-based 95% intervals.

```{r, eval = FALSE}
# plot average-case predicted probabilities
ggplot(
  average_output,
  aes(x = ________, y = __________)
) +
# create confidence intervals
  geom_ribbon(
    aes(
      ymin = ______,
      ymax = ______
    ),
    alpha = 0.25
  ) +
  geom_line() +
  theme_bw() +
  labs(
    x = "Same-Party Presidential Vote",
    y = "Predicted Probability of Discussing Immigration"
  )
```

**Question:** For the hypothetical candidate we constructed, what happens to the predicted probability of discussing immigration as the district becomes more favorable to the candidate's party?

*Response:* 

**Question:** Who exactly does this curve describe? Which values in our hypothetical candidate did we choose ourselves?

*Response:* 

This second question points to an important limitation of the average-case approach. We get an intuitive prediction for one clearly defined hypothetical candidate, but the marginal change in predicted probability can depend on which hypothetical candidate we choose. For example, if we were comparing among Republican incumbent candidates, there might be a different marginal effect for `district_lean` on the probability of discussing immigration.

## Observed-Case Approach

An alternative is the **observed-case approach**.

Instead of constructing one hypothetical candidate and holding the other predictors at selected values, we'll retain the actual values of the other covariates for every candidate in our estimation sample.

For each value of `district_lean`, we'll:

1. Take every candidate in the estimation sample.
2. Set `district_lean` equal to that value for everyone.
3. Leave candidate type, fundraising, party, primary competition, and gender at their observed values.
4. Calculate a predicted probability for every candidate.
5. Average those predicted probabilities across candidates.
6. Repeat this process for every simulated set of coefficients.

This gives us the *average predicted probability across the candidates represented in our estimation sample* at each hypothetical value of `district_lean`.

First, pull the original design matrix from our fitted model.

```{r}
# create design matrix using observed covariate values
observed_matrix <- model.________(model_logit)

# inspect dimensions
dim(observed_matrix)
```

We'll create an empty matrix to store our simulated average predicted probabilities. Each row will represent one coefficient simulation and each column will represent one value of `district_lean`.

```{r}
# initialize matrix for observed-case simulation results
observed_sims <- matrix(
  NA,
  nrow = n_sim,
  ncol = length(district_vec)
)
```

Now loop through the different hypothetical values of `district_lean`.

```{r}
# calculate observed-case predicted probabilities
for (j in seq_along(district_vec)) {

  # make temporary copy of observed design matrix
  temporary <- observed_matrix

  # replace district lean for ALL candidates
  # with current hypothetical value
  temporary[, "district_lean"] <- district_vec[j]

  # calculate predicted probabilities for every candidate
  # under every simulated set of coefficients
  pp <- plogis(
    temporary %*% t(beta_sims)
  )

  # average across candidates for each coefficient simulation
  observed_sims[, j] <- colMeans(pp)
}
```

There are several steps happening inside this loop, so let's unpack them.

At each value of `district_lean`, every candidate is temporarily assigned the same hypothetical district lean. Their other characteristics remain exactly as observed. We then calculate a predicted probability for every candidate under every simulated coefficient vector and average those predictions across candidates.

Importantly, we take *one additional average* in the observed-case approach. Rather than calculating the predicted probability for one hypothetical candidate, we first calculate predictions for all of the observed candidates and then average over those predictions.

Now summarize the simulated averages.

```{r}
# calculate point estimates and simulation intervals
observed_output <- data.frame(
  district_lean = district_vec,
# average predicted probability across simulations
  pp_avg = apply(
    ______,
    2,
    mean
  ),
# lower bound of 95% confidence interval
  lb = apply(
    ________,
    2,
    quantile,
    probs = 0.025
  ),
# lower bound of 95% confidence interval
  ub = apply(
    ______,
    2,
    quantile,
    probs = 0.975
  )
)

# inspect output
head(observed_output)
```

Plot the observed-case results.

```{r}
# plot observed-case predicted probabilities
ggplot(
  observed_output,
  aes(x = ______, y = ________)
) +
  geom_ribbon(
    aes(
      ymin = ________,
      ymax = ________
    ),
    alpha = 0.25
  ) +
  geom_line() +
  theme_bw() +
  labs(
    x = "Same-Party Presidential Vote",
    y = "Average Predicted Probability of Discussing Immigration"
  )
```

**Question:** In the observed-case approach, which variable did we change and which variables did we leave at their observed values?

*Response:* 

**Question:** What are we averaging over at each value of `district_lean`?

*Response:* 


## Comparing Average-Case and Observed-Case Predictions

Finally, let's put the two sets of predictions together.

```{r}
# label each approach
average_output$approach <- "Average Case"
observed_output$approach <- "Observed Case"

# combine dataframes
comparison_output <- bind_rows(
  average_output,
  observed_output
)

# compare point estimates from both approaches
ggplot(
  comparison_output,
  aes(
    x = district_lean,
    y = pp_avg,
    linetype = approach
  )
) +
  geom_line(
    linewidth = 1
  ) +
  theme_bw() +
  labs(
    x = "Same-Party Presidential Vote",
    y = "Predicted Probability of Discussing Immigration",
    linetype = "Approach"
  )
```

The two lines should follow a broadly similar substantive pattern, but they are not identical. This is not a coding error!

The average-case approach asks about one hypothetical candidate whose other characteristics we selected. The observed-case approach asks what the average predicted probability would be across the candidates represented in our estimation sample while changing only `district_lean`.

Because the logit model is nonlinear, calculating a prediction for a candidate with selected or "average" characteristics is not generally the same thing as calculating predictions for all of the observed candidates and then averaging those predictions.

**Question:** Why aren't the average-case and observed-case predictions identical?

*Response:* 

**Question:** Suppose your substantive goal were to describe how predicted immigration discussion changes across district lean for the population of candidates represented in this dataset. Which approach would you prefer, and why?

*Response:* 


## Evaluating Model Discrimination

So far, we've focused on *interpreting* our model. We've asked how the predicted probability of discussing immigration changes as `district_lean` changes.

We can also ask a different question: *how well does the model distinguish between candidates who actually discuss immigration and candidates who do not?*

Our model gives every candidate a predicted probability between 0 and 1. We could choose a threshold—for example, 0.50—and classify candidates with predicted probabilities above that threshold as discussing immigration. But 0.50 is ultimately an arbitrary cutoff. A different threshold would produce a different combination of correctly and incorrectly classified observations.

The **Receiver Operating Characteristic (ROC) curve** evaluates this tradeoff across all possible classification thresholds.

For each possible threshold, the ROC curve compares:

- the *true-positive rate*, or the proportion of observed 1s correctly identified as 1s; and
- the *false-positive rate*, or the proportion of observed 0s incorrectly identified as 1s.

A model that does a better job separating the two groups will produce an ROC curve that bows farther toward the upper-left corner of the plot. A model with no discriminatory ability will fall along the 45-degree diagonal.

### ROC Curve

First, we'll generate the ROC curve using the fitted probabilities from our model. The `roc()` function needs two main pieces of information:

- `predictions` is a numeric vector containing the model's predicted probability for each observation. Here, we use predict(model_logit, type = "response"), which returns the fitted probability that each candidate discusses immigration.
- `labels` contains the actual observed outcome for each of those same observations. In our case, these are the observed values of immigration, coded 0 or 1.

Importantly, these two vectors line up observation by observation. For each candidate, predictions tells us what probability the model assigned to discussing immigration, while labels tells us whether that candidate actually discussed immigration.

```{r}
# create ROC curve
model_roc <- ______(
  predictions = ______(
    ______,
    ________ = "______"
  ),
  labels = as.factor(
    model.response(
      model.frame(model_logit)
    )
  )
)

# plot ROC curve
______(________)

```

**Question:** What would we expect the ROC curve to look like if our model were no better than random at distinguishing candidates who discuss immigration from candidates who do not?

*Response:* 


### Area Under the ROC Curve

Rather than interpreting the entire ROC curve visually, we can summarize its discriminatory ability using the **area under the ROC curve (AUROC)**. You will also commonly see this referred to simply as the **AUC**.

```{r}
# calculate AUROC ('auc' command)
______(______)
```

AUROC ranges from 0 to 1. A value of 0.5 indicates no discriminatory ability, while a value of 1 indicates perfect discrimination. Values below 0.5 indicate that the model is systematically ranking the two groups in the wrong direction.

A useful substantive interpretation is to imagine randomly selecting one candidate who discusses immigration and one candidate who does not. The AUROC is the probability that the model assigns the candidate who discusses immigration the *higher predicted probability*.

**Question:** What is the AUROC for our model? Interpret that value in the context of candidates discussing immigration.

*Response:* 

*Note:* The ROC curve and AUROC above are calculated using the *same observations used to estimate the model*. They therefore describe the model's *in-sample discrimination*. They do not tell us how well the model would perform on entirely new candidates that were not used to estimate it. Evaluating out-of-sample performance requires a separate test sample or a procedure such as cross-validation.

AUROC also evaluates *discrimination*, not whether the predicted probabilities are perfectly calibrated. A model can rank 1s above 0s relatively well without every predicted probability matching the corresponding observed frequency.

