---
title: "POLI784 SP26 Lab - Week 2: Programming concepts & best practices"
author: "Sayman J. Rohleder-Stribl"
date: today
format: html
---

```{r setup, include=FALSE}
# set output options for rendered file
knitr::opts_chunk$set(
  eval = TRUE, # evaluate code in rendered output
  echo = TRUE, # show code in rendered output
  include = TRUE, # include output in rendered output
  message = FALSE, # hide messages in rendered output
  warning = FALSE, # hide warnings in rendered output
  error = FALSE, # hide errors in rendered output
  cache = FALSE, # do not cache results between renders
  tidy = FALSE, # do not reformat code
  collapse = TRUE, # collapse code and console output in rendered output
  comment = "#>>>" # prefix for console output lines in rendered output
)
```

## RESOURCES FOR RSTUDIO INLINE COMPLETIONS

- `Posit` Github Copilot guide: <https://docs.posit.co/ide/user/ide/guide/tools/copilot.html>

## Sources

Source materials for this lab include:

- `R` documentation: <https://cran.r-project.org/>
- `Posit` website: <https://posit.co/>
- `RStudio` documentation: <https://posit.co/download/rstudio-desktop/>
- `Positron` documentation: <https://positron.posit.co/>
- `R Markdown` documentation: <https://rmarkdown.rstudio.com/>
- `Quarto` documentation: <https://quarto.org/>
- `RStudio` projects documentation: <https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects>
- `GitHub` documentation: <https://docs.github.com/en/get-started/quickstart>
- `git` documentation: <https://git-scm.com/>
- `renv` documentation: <https://rstudio.github.io/renv/articles/renv.html>, <https://posit.co/blog/renv-project-environments-for-r/>
- `tidyverse` style guide and linters: <https://style.tidyverse.org/>, <https://styler.r-lib.org/>, <https://github.com/r-lib/lintr>

## Introduction to labs

Welcome! The lab portion of POLI784 is designed to provide you with hands-on experience implementing the concepts and methods you learn about in the lecture portion of the class. We'll be doing this using the `R` programming language. To get you up and running for the semester, this week's lab will focus on introducing you to basic concepts and tools that you will encounter throughout the semester.

While this week will predominantly be structured like a guide / tutorial, future labs will revolve mostly around brief, structured exercises for you to solve. These exercises are mainly designed to give you some practice before having to tackle more involved statistical programming assignments on your problem sets. While lab attendance is mandatory, you are *not* required to complete the lab exercises or submit them, as they will not be graded. You are, however, strongly encouraged to complete them and -- if you find the time to do so -- submit them before the following Tuesday's lecture session on the lab-specific Canvas page to ensure that you get sufficient practice before solving the p-sets. This will also allow you to get feedback from me on aspects of the labs that you may not have been able to solve or want to know more about.

## A word on p-sets, office hours, and other logistics

As you know, you will have to solve six graded problem sets for this course. Santiago has told you most of what you need to know about them already, but we have decided to add one additional element to support you in your learning this semester -- an oral check-in after each p-set. While you'll be submitting your p-sets in groups, the check-in will be a one-on-one with me. I am not yet completely set on how to do this logistically, but you'll likely be signing up for ~10min office hour slots after each p-set. These check-ins will be rather informal and ungraded, and their purpose is two-fold. First, you'll get a chance to discuss your solutions for the p-set and any challenges you faced while working on it. Essentially, you'll be able to get immediate feedback from me on how to solve issues that you may have encountered. Second, it is an incentive to not free-ride on the assignments, as having these check-ins will allow us to hold each individual group member to account when it comes to explaining the solutions submitted.

When working on the p-sets you should collaborate closely with your group members (and other students in the class, if desired). This will hopefully provide you with a chance to get help from and provide help to others, which should make solving the p-sets easier. At the same time, make sure to make ample use of office hours whenever you feel stuck on the assignments.

With that being said, please take the time to review the information below before solving the (very brief) exercise at the end of the file. Once you have read and worked through this file, you should be equipped with the basic knowledge and tools to tackle the programming tasks you will be presented with throughout the rest of the semester. At the same time, computers and software -- despite being deterministic beasts -- can sometimes be uncooperative ... we've all been there! So, if you find yourself struggling with anything at all related to the labs, please do not hesitate to reach out to me ASAP. A 5-minute debugging session during office hours can save you literal *hours* of time. The same goes for discussing the substantive content of the lecture itself too, of course. Make use of my (and Santiago's) office hours regularly -- it will make this semester a lot more manageable.

## Programming languages, packages, and IDEs

When using `R` (or any other programming language, for that matter) to program, you will be engaging with related but distinct parts of the programming ecosystem. It is helpful to know how these parts relate to and differ from one another, as this will make it easier for you to problem solve common issues that may arise down the road. The three main parts of the programming ecosystem that I think are worth being aware of for this course are:

- the programming language itself (in this case, `R`)
- modules that extend the functionality of the core programming language (in the case of `R`, these are called 'libraries' or 'packages' and are hosted on repositories such as `CRAN` or `GitHub`)
- the software environment (i.e., *integrated development environment* or *IDE*) you use to write, edit, and execute code, as well as view outputs (in the case of `R`, generally `RStudio`)

Let's start with the language. A programming language (or really the symbiosis between the language and its associated interpreter, but let's not get in the weeds here ...) is the heart of the programming operation. Just like the engine -- rather than the chassis or interior design elements -- propels a car forward, the programming language allows your computer to execute your instructions. We tend to think relatively little about how our code gets executed mechanically because the process is hidden from view (like the engine of a car!), which is why when issues arise, they tend to be hard for us to solve. Fortunately, `R` is generally pretty user-friendly, so once you have set it up on your computer, it will hopefully give you relatively little trouble. To get `R` up and running on your system, download and install it from [CRAN](https://cran.r-project.org/).

While `R` is a rather versatile language, the people who created and maintain the core distribution of `R` cannot write code to cover all possible use cases that we care about. Hence, modules -- often referred to as packages or libraries -- exist. Packages contain code written in the `R` language that expands the functionality of the core language. We will be using many packages that provide us with off-the-shelf software implementations for running statistical models of interest, which is nice because it means that we do not have to implement from scratch any of the fancy models you will learn about in this class! To gain access to the functionality of a package, you will generally download it from `CRAN` and install it using the below workflow.

```{r install-package, eval=FALSE}
# install the 'dplyr' package from CRAN
install.packages("dplyr")

# load package into session to be able to access its functions
library(dplyr)
```

Finally, while it is possible to execute R directly from the command line / terminal, few are self-loathing enough to do so. As is the case with most programming languages, `R` is best utilized in conjunction with an IDE. An IDE is a software environment that provides a graphical user interface that you can use to open, view, edit, and execute scripts, see the objects that you have created in your session, and view output of your code, etc. Since `R` is a programming language built from the ground up with data analysis and statistical programming in mind, it has its own tailor-made IDE designed by [Posit](https://posit.co/) (formerly the `RStudio Project` -- an organization which has created a whole ecosystem of data analysis tools, including for `R`) for just that purpose: `RStudio`. You can download and install `RStudio` from the [Posit website](https://posit.co/download/rstudio-desktop/).

If you have used `R` in conjunction with `RStudio` before, I would recommend you stick with `RStudio` because it provides great functionality and grappling with a new IDE would add additional overhead to your learning process this semester. I do want to say, however, that `Posit` has recently released the `Positron` IDE, which is a third-party fork of `VSCode` (once again [the most popular IDE in the world in 2025](https://survey.stackoverflow.co/2025/technology#1-dev-id-es), and it's not even close!) that unites the best features of `RStudio` for data analysis and `VSCode` for general-purpose programming in a polyglot IDE. I personally prefer `Positron` to `RStudio` (not least because it natively supports `Python`), so for those of you like me who are fond of `VSCode`, give `Positron` a try! Regardless of your choice, the underlying functionality of `R` remains the same, so this comes down to personal preference.

Of course, like any other software, the above three elements have to be maintained by updating them. To update `R` itself, you can use the `installr` package. Simply run the below code to check for updates and, if some are available, install them by following the pop-up instructions. I would recommend doing this every other week or so, just to make sure you are running the latest version of `R`.

```{r update-r, eval=FALSE}
# install installr package, if not already present
if(!require("installr")) install.packages("installr")

# check for updates
installr::updateR() # NOTE: the package_name::function_name() syntax calls a function without loading its (already installed) package into the session first
```

Once you have installed a new version of `R`, `RStudio` will likely be uncooperative and keep using the old version of `R` (unless you uninstalled the old version during the updating process, which I would *not* recommend doing). To fix this, make sure to press and hold the `Ctrl` key (or `Cmd` key on Mac) while starting `RStudio` (i.e., right after clicking on the `RStudio` symbol). This will prompt `RStudio` to ask you which version of `R` you want to use. Select the newly installed version once, and `RStudio` will remember to use this version for future sessions. At the same time, should you ever want to revert back to an older version still present on your local machine (e.g., because a specific project of yours requires said older version for reproducibility, etc.), you can do so by following the same procedure.

Finally, packages and IDEs have to be updated periodically too, and I would recommend doing this every other week as well. To update packages, it is easiest to use `RStudio`'s GUI by selecting *Tools* > *Check for Package Updates...* and then following the instructions, if there are packages that need updating. Similarly, by selecting *Help* > *Check for Updates...* you can check whether your version of `RStudio` needs updating.

You might feel like updating `R`, packages, and `RStudio` is simply busy work, but especially in the case of newly released packages (and we will use a few of those during the semester!) and most IDEs, you can gain *a lot* of functionality from one update to another! To illustrate this point, imagine you hadn't updated your `RStudio` since late 2022 -- that would mean that you would have no access to inline completion models (more on those below), which increase your productivity drastically! More importantly, packages tend to depend on `R` being up-to-date to function properly, so you generally want to avoid being too far behind on the updating schedule, as that will cause dependency issues.

## Typesetting and markup languages

Now that we have set up the tools you need to program, it's time to talk about the tools you can use to document your code and communicate your results. For doing your own research I recommend using a combination of scripts (e.g., `.R` files for `R` code, `.py` files for `Python` code, etc.) for your code routines and typesetting files (e.g., `.tex` files for `LaTeX`) for writing up your results. This separation of analysis and write-up is actually quite helpful, in my view, because it allows you to run tons of code while only incorporating the results you actually need into the final manuscript, leaving the rest of the analysis code 'behind the scenes'.

For the labs and p-sets in this class, however, it is quite helpful to use files that allow you to combine code and written text in a single place. Those of you who have been using `R` will likely also have used [R Markdown](https://rmarkdown.rstudio.com/) files before. `R Markdown` combines the markup language functionality of `Markdown` (which allows you to format text using simple syntax) with the ability to embed `R` (and other programming language) code chunks that can be executed directly from within the document. The combined text, code, and code output can then be rendered / compiled (i.e., *knit*, in the `R Markdown` lingo) into a nicely formatted document (e.g., `HTML`, `PDF`, `Word`, etc.).

I am happy for you to use `R Markdown` for your problem sets, but I will be using [Quarto](https://quarto.org/) instead. Don't fret -- `Quarto` is simply the next-generation, language-agnostic version of `R Markdown`, and the two are functionally de facto identical to one another (as they are both once again products of the `Posit` crew). You can use the `Quarto` files provided for the labs the same way that you would an `.Rmd` file, since `Quarto` and `RStudio` are compatible (although other IDEs and `.Rmd` files are not, which is why `Quarto` exists!). The only difference will be that `RStudio` will show a *Render* button instead of a *Knit* button above your file.

## Projects and virtual environments

While you won't get around using `R`, packages, some IDE, and some markdown or typesetting language to complete this course, following the advice in the remaining sections is technically optional. However, I believe that following the best practices described in this and the following sections will *drastically* improve your quality of life -- especially in the long term! (Just like a financial investment compounds over time, so does investing a few hours at the start of your academic career via saving yourself hundreds of hours down the road!)

The first best practice is to use `RStudio` [projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) (`Positron` projects also exist and are much easier to use because they require no setup!). The description of how to create and use a project at the link provided above should cover your bases, so I will not go through an example here. But should you have questions or need guidance, let me know and I'll help you set up things! Among many advantages, using projects ensures that your working directory is set to the folder within which you initialized the project and that your source documents and session settings are restored to where you left off last time. This is convenient because it allows you to switch between projects located in entirely different folders with entirely different files open in, for example, `RStudio` with a single click, rather than having to manually set the working directory and re-open files each time you want to switch contexts.

The real power of projects, however, is unlocked when combining them with [GitHub](https://docs.github.com/en/get-started/quickstart) repos (or just `git` projects) and / or virtual environments. While version control using `git` and `GitHub` goes beyond the scope of this lab and the lecture, I highly recommend that you use `Posit`'s [virtual environments](https://posit.co/blog/renv-project-environments-for-r/) for `R` via the [renv](https://rstudio.github.io/renv/articles/renv.html) package. Combining projects with virtual environments has several key advantages:

- `R` packages are installed into project-specific libraries, which is especially helpful when you need to use two different versions of the same package for different projects or when packages from different projects are not compatible with one another (which is sometimes the case with experimental packages, like many early-stage statistical modeling packages!)
- sharing the project-specific library is largely automated, meaning that your coauthors (or you on a different computer) can get up and running with all project dependencies with minimal effort 
- your project results are easily reproducible because the state of your packages at the time of running the analysis can be captured and recreated in the future without any manual work

Considering the above, here's what I would recommend you do before starting each lab assignment or problem set for this class:

- create a new `RStudio` project in an empty folder (e.g., a subdirectory of the POLI784 folder on your local machine) dedicated to the specific lab or problem set (NOTE: make sure that there is no existing `.Rproj` file in a parent or child directory of the folder)
- initialize a `renv` virtual environment for the project by running the code provided below
- once initialized, the virtual environment will automatically activate whenever you open the project in your IDE
- install any packages you need for the project (packages will be installed into the `renv`-specific library, rather than the global `R` library) using ...
  - ... either `renv::restore()` to install the packages saved in the `renv.lock` file provided for each lab
  - ... or `install.packages()` if you want to install them one-by-one at the top of your file, as usual 
- after installing all required packages, snapshot the state of the project library by running `renv::snapshot()`
- repeat for each lab assignment or problem set, ensuring that projects stay siloed from one another and your main `R` installation stays clean

```{r renv-init, eval=FALSE}
# install package, if not already present
if(!require("renv")) install.packages("renv")

# initialize renv for current project
renv::init()
```

## Style and linting

I am (unreasonably) obsessed with code style and linting (what can I say, I'm Austrian ... we like order ¯\\\_(ツ)_/¯), but I will try to keep this section to a minimum. Style refers to the way code is formatted and structured. While there are no hard rules for how code *has* to be styled, writing badly formatted code is the same to me as writing an essay with bad grammar, sentence structure, and punctuation: It might run (get the point across), but using more conventional style would make it easier to read and maintain (understand). While some languages have very clear, community-wide style guides (e.g., [PEP 8](https://peps.python.org/pep-0008/) for `Python`), the `R` community tends to play a lot faster and looser -- but fortunately, the good folks at `Posit` have come to the rescue with their [tidyverse style guide](https://style.tidyverse.org/).

I am not going to force you to use a specific style (or any coherent style at all, for that matter), but I think it is worth considering sticking to the `tidyverse` style guide for `R` code linked above because it will make your code easier to read and maintain, which will in return hopefully increase your desire to work on your code! At the same time, thinking about *how* to write code should not slow down your process of actually *writing* code, so there's a trade-off there. Fortunately, linters (like `Posit`'s [styler](https://styler.r-lib.org/) and [lintr](https://github.com/r-lib/lintr)) are a thing -- and you can even use them to format code via the click of a button in your `RStudio` GUI! Linters are pre-written scripts distributed as packages that you can use to automatically adjust your code to fit a given style. Importantly, automatically linting a file does not change code *functionally*, which means that you do not have to worry about it breaking your script. This way you can write code, automatically adjust it to match good style, and hopefully learn how to write nicely structured code in the process -- all without having to actively focus on it too much!

To illustrate these points, here's an example of some badly styled code. Install the `styler` package and use the `styler::style_file()` function to automatically adjust the code to match the `tidyverse` style guide. Pass the path to this `.qmd` file as an argument to the function and see what happens to the code chunk below! It's easiest to execute `style_file()` directly in console, rather than adding it to a script and executing it from there.

```{r bad-style-example}


# badly styled function definition
badly_styled_function <-function(x,y){result<-x+y;if(result>10){return(TRUE)}else{return(FALSE)}}

# badly styled function call
data <- data.frame(age = numeric(0),
  height = numeric(0),
  weight = numeric(0))

```

## Pseudo code and commenting

Finally, let's talk about comments. Comments have always served a dual purpose in programming. On the one hand, comments are used to explain a given code chunk to the human reader (who may, in fact, be you in the future!). On the other hand, comments can also be used to scaffold your code before you write it. The latter is referred to as pseudo code and can be understood as instructions (i.e., an algorithm) written out in natural language (e.g., English). Here's an example (although I wouldn't recommend using a loop for this specific task -- see discussion at the end of the file!):

```{r pseudo-code-example, eval=FALSE}

# loop over rows
...

  # grab value in column A
  ...

  # add 5 to value
  ...

  # store new value in column B
  ...

```

As should be evident, once you have written the code to match the pseudo code above, these comments can be kept as documentation -- two birds, one stone! The irony of the whole thing is that when I learned to program, instructors rightfully used to say that 'Pseudo code is the human-readable form of code, since only code written in a programming language can be interpreted and executed by a computer'. And that brings us to the third purpose of commenting / pseudo code: Instructing LLMs to write clean, functional code for you!

Nowadays, computers *can* (in a sense) interpret natural language, with LLMs taking your natural language instructions and turning them into pretty much whatever output you like (although a lot of that hinges on tool calls, which means that there is actually still code being executed under the hood for that to work). This means that you can now use pseudo code not only to plan your approach, but also to get an inline completion model (like those available to you in `RStudio` and other IDEs) to write the actual machine-readable `R` code for you -- what a time to be alive!

It might seem absurd to do this when you could also just write a brief prompt and get an LLM to write hundreds of lines of code for you in one go, but especially at the level of programming proficiency that tends to be common among 1st-year grad students, that approach often ends in tragedy. Getting large-scale output from an LLM requires the skills to properly *validate* that output, and that's precisely where things tend to go wrong. By writing pseudo code and getting code completions a few lines at a time, you are necessarily actively involved in reviewing and adapting the code (human in the loop!), which will not only help you spot mistakes but also force you to actively plan out your problem solving approach and closely engage with the output. This will help you learn from the LLM and teach you how to guide it, rather than letting it rob you of an opportunity to learn by letting it guide you.

Let's complete an example of how to query an inline completion model to write code for you. For each of the steps listed below, write one or several comments and accept the completion suggested by your IDE's inline code completion model. (You can copy and paste the bullet points with a '#' in front into the code chunk and use them as comments, if you want). If necessary, edit the code manually to ensure that it works as intended. By writing these comments, you are effectively writing pseudo code to think through your implementation, while simultaneously writing comments that serve as documentation for your code.

- set seed for reproducibility
- start timer to measure computational efficiency of code
- create an empty dataframe with age, height, and weight columns
- populate dataframe with simulated data for 10,000 individuals
  - sample age between 18 and 80
  - sample height from normal distribution with mean 170 and sd 10
  - sample weight from normal distribution with mean 70 and sd 15
- print the values of all three variables for row 333 to console
- stop timer and print elapsed time

Once you have the code, lint it using `styler::style_file()` in console and see if the style of the code changes.

```{r inline-completion}

```

This was, of course, a trivial task -- and the inline completion model should have filled in pretty much everything from the get go because it saw both my instructions above and its previous output down below. But I hope it illustrates the point that writing LLM-generated code using comments is really fast, while also making you engage directly with the output without having to ever switch context. For a more realistic application, feel free to create a new code chunk, think of anything at all that you want to code up, and code it in its entirety by creating one comment at a time and accepting the associated inline completions. Give it a try -- see if you can get vibe coding to do the heavy lifting for you.

Seeing the output that the inline model gave me, there is a second lesson to be learned here: Vectorization is generally preferred to looping. While loops are extremely powerful and versatile (which is why we will be using them quite a bit when simulating), they tend to be overused. While this is not inherently a problem, looping over rows (or columns) and performing computations one-by-one can be slow, depending on the context (as is the case above!). Hence, it is better to use a vectorized operation (defined as code operating on a whole vector of values at once, rather than each value of that vector one-by-one), if possible. Here's an example of how to vectorize the above code chunk (obtained simply via the comment 'vectorize the above code chunk' and accepting and slightly editing the completions suggested by the inline model for each line!). Run both chunks and see the difference in the time it takes your computer to perform the computations. Have a look at what has changed from chunk one to chunk two.

**ORIGINAL OUTPUT:**

```{r inline-completion-example}
# set seed for reproducibility
set.seed(333)

# start timer to measure computational efficiency of code
start_time <- Sys.time()

# create an empty dataframe with age, height, and weight columns
data <- data.frame(age = numeric(0), height = numeric(0), weight = numeric(0))

# simulate data for 100k individuals
for (i in 1:10000) {
  age <- sample(18:80, 1)
  height <- rnorm(1, mean = 170, sd = 10)
  weight <- rnorm(1, mean = 70, sd = 15)
  data <- rbind(data, data.frame(age = age, height = height, weight = weight))
}

# print the values for row 333 to console
print(data[333, ])

# stop timer and print elapsed time
end_time <- Sys.time()
elapsed_time <- end_time - start_time
print(elapsed_time)
```

**VECTORIZED OUTPUT:**

```{r vectorized-code}
# set seed
set.seed(333)

# start timer
start_time <- Sys.time()

# simulate data for 10,000 individuals
N <- 1e4
data <- data.frame(
  age = sample(18:80, N, replace = TRUE),
  height = rnorm(N, mean = 170, sd = 10),
  weight = rnorm(N, mean = 70, sd = 15)
)

# print the values for row 333 to console
print(data[333, ])

# end timer and print elapsed time
end_time <- Sys.time()
elapsed_time <- end_time - start_time
print(elapsed_time)
```

As you can see, the time elapsed is drastically reduced by using a vectorized operation compared to a for-loop!
