---
title: "Lab 5: Count, Ordered, and Multinomial Models"
author: "Audrey Sachleben"
format: html
date: "September 18, 2026"
---

```{r setup, include=FALSE}
# set output options for rendered file
knitr::opts_chunk$set(
  eval = TRUE, # evaluate code in rendered output
  echo = TRUE, # show code in rendered output
  include = TRUE, # include output in rendered output
  message = FALSE, # hide messages in rendered output
  warning = FALSE, # hide warnings in rendered output
  error = FALSE, # hide errors in rendered output
  cache = FALSE, # do not cache results between renders
  tidy = FALSE, # do not reformat code
  collapse = TRUE, # collapse code and console output in rendered output
  comment = "#>>>" # prefix for console output lines in rendered output
)
```

```{r, include = FALSE}
rm(list = ls())

# load packages (install if necessary)
library(tidyverse)
library(haven)
#install.packages("AER")
library(AER)
#install.packages("dataverse")
library(dataverse)
#install.packages("ordinal")
library(ordinal)
library(nnet)
library(MASS)
```

## Overview

So far, we have focused on binary outcome variables. Today, we will extend that framework to three additional kinds of outcomes: count outcomes, ordered categorial outcomes, and multinomial outcomes.

Across each of these model types, we'll estimate the appropriate model and see what information can be gathered from it. Then, we'll generate and visualize (i.e., plot) predicted outcomes or predicted probabilities related to our predictors of interest. *Note:* the process we'll focus on today is something that you'll come back to time and time again in your own research, especially when thinking about how to best convey your results in a paper.

By the end of the lab, you should be able to fit and interpret Poisson/negative binomial, ordered logit, and multinomial logit models and generate substantively interpretable predictions from each.

## Count Outcomes

We will begin with a count outcome. Count variables are discrete, bounded below at zero, and can only take non-negative integer values. Examples include the number of protests in a country-year, the number of bills sponsored by a legislator, or the number of executive orders issued by a governor.

For this section, we will use replication data from Barber, Bolton, and Thrower (2019) on executive unilateralism under divided government. Our dependent variable, `num_eos`, is the number of executive orders issued by a governor in a given month.

We will use five predictors:

- `divided_gov1`: whether the governor serves under divided (1) or unified (0) government
- `veto_proof`: whether there is a veto-proof majority in the state legislature
- `house_polarization`: polarization in the lower chamber of the state legislature
- `governor_party_pres_vote`: presidential vote share for the governor's party
- `governor_power`: an index of gubernatorial power

First, read in the data and examine the distribution of the dependent variable.

```{r}
# Pull the replication data from Dataverse
executives <- read_csv("executives_data.csv")

# examine distribution of data
executives %>%
  ggplot(aes(x = num_eos)) +
  geom_histogram(binwidth = 5) +
  theme_bw() +
  labs(
    x = "Number of Executive Orders",
    y = "Frequency"
  )
```

**Question: What features of this variable make OLS a potentially poor choice? In particular, think about the possible values of the dependent variable and the shape of its distribution.**

### Poisson Regression

A standard starting point for count outcomes is the Poisson regression model. The Poisson model uses a log link, so the conditional mean of the outcome is constrained to be positive:

$$
\log(\lambda_i) = X_i\beta,
$$

or equivalently,

$$
\lambda_i = \exp(X_i\beta),
$$

where $\lambda_i = E(Y_i\mid X_i)$.

Fit a Poisson regression predicting the number of executive orders.

```{r}
# fit a poisson model
poisson_model <- glm(
  num_eos 
  ~ divided_gov1 
  + veto_proof 
  + house_polarization 
  + governor_party_pres_vote 
  + governor_power,
  data = executives,
  family = poisson(link = "log")
)

#inspect results
summary(poisson_model)
```

Because the coefficients are expressed on the log-count scale, exponentiating a coefficient gives its multiplicative effect on the expected count.

```{r}
# exponentiate coefficients
exp(coef(poisson_model))
```

**Question: Select one coefficient and interpret its exponentiated value.**

A key assumption of the Poisson model is equidispersion:

$$
E(Y_i\mid X_i) = Var(Y_i\mid X_i).
$$

In many political science applications, the variance is substantially larger than the mean. This is called *overdispersion*.

We can formally evaluate overdispersion using `dispersiontest()` from the `AER` package.

```{r}
# test for overdispersion
dispersiontest(poisson_model)
```

The null hypothesis is that the dispersion parameter equals 1, as assumed by the Poisson model. Here, the estimated dispersion parameter is approximately 83.6, meaning that the variance is substantially larger than the mean. The small p-value ($p < 0.001$) provides strong evidence of overdispersion, suggesting that the Poisson model is not appropriate for these data. In this case, it therefore likely makes sense to pivot to a negative binomial model.

### Negative Binomial Regression

The negative binomial model relaxes the Poisson assumption that the conditional mean and variance are equal. It is therefore often useful when the data are overdispersed.

Fit the same model using `glm.nb()` from the `MASS` package.

```{r}
# fit a negative binomial model
nb_model <- glm.nb(
  num_eos 
  ~ divided_gov1 
  + veto_proof 
  + house_polarization 
  + governor_party_pres_vote 
  + governor_power,
  data = executives
)

# inspect results
summary(nb_model)
```

**Question: Compare the Poisson and negative binomial results. Do the coefficient estimates change substantially? What happens to the standard errors? Why might ignoring overdispersion affect our statistical inferences?**

### Predicted Counts

The coefficients from either model are not directly expressed in numbers of executive orders. We can instead generate predictions on the response scale.

We will examine how predicted executive orders change across values of gubernatorial power under unified and divided government. We will hold the remaining covariates at their sample means.

```{r}
# generate grid for hypothetical observations
count_grid <- expand_grid(
  governor_power = seq(
    min(executives$governor_power, na.rm = TRUE),
    max(executives$governor_power, na.rm = TRUE),
    length.out = 100
  ),
  divided_gov1 = c(0, 1)
) %>%
  mutate(
    veto_proof = mean(executives$veto_proof, na.rm = TRUE),
    house_polarization = mean(executives$house_polarization, na.rm = TRUE),
    governor_party_pres_vote = mean(executives$governor_party_pres_vote, na.rm = TRUE)
  )

# generate predictions and standard errors on the link scale
count_predictions <- predict(
  nb_model,
  newdata = count_grid,
  type = "link",
  se.fit = TRUE
)

# convert predictions and 95% confidence intervals back to expected count scale
count_grid <- count_grid %>%
  mutate(
    predicted_eos = exp(count_predictions$fit),
    conf_low = exp(
      count_predictions$fit -
        qnorm(0.975) * count_predictions$se.fit
    ),
    conf_high = exp(
      count_predictions$fit +
        qnorm(0.975) * count_predictions$se.fit
    )
  )

# plot predictions with 95% confidence intervals
# plot should show the number of executive orders as a function of gubernatorial power
# split out by unified and divided government
count_grid %>%
  mutate(
    divided_gov = factor(
      divided_gov1,
      levels = c(0, 1),
      labels = c(
        "Unified government",
        "Divided government"
      )
    )
  ) %>%
  ggplot(aes(
    x = governor_power,
    y = predicted_eos,
    color = divided_gov,
    fill = divided_gov,
    linetype = divided_gov
  )) +
  geom_ribbon(
    aes(
      ymin = conf_low,
      ymax = conf_high,
      group = divided_gov
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  scale_color_manual(
    values = c(
      "Unified government" = "#3B6EA8",
      "Divided government" = "#B35C44"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Unified government" = "#3B6EA8",
      "Divided government" = "#B35C44"
    )
  ) +
  theme_bw() +
  labs(
    x = "Gubernatorial Power",
    y = "Predicted Number of Executive Orders",
    color = NULL,
    fill = NULL,
    linetype = NULL
  )
```

**Question: Describe the substantive relationship shown in the plot. Why is this easier to interpret than the raw negative binomial coefficients?**

## Ordered Outcomes

An ordered dependent variable contains categories with a meaningful ranking, but we cannot assume that the distance between adjacent categories is constant.

For example, consider agreement with the statement:

> "A working mother can establish just as warm and secure of a relationship with her child as a mother who does not work."

Respondents can answer:

1. Strongly disagree
2. Disagree
3. Agree
4. Strongly agree

These categories have a clear ordering, but the difference between "Strongly disagree" and "Disagree" is not necessarily the same as the difference between "Agree" and "Strongly agree."

We will use data from the 1977 and 1989 General Social Survey. The predictors are:

- `male`: respondent is male
- `white`: respondent is white
- `age`: respondent age
- `ed`: years of education
- `prst`: occupational prestige

Read the data and prepare it for analysis. 

```{r}
# load data
hw <- read_csv(
  url("http://faculty.washington.edu/cadolph/mle/ordwarm2.csv"),
  show_col_types = FALSE
)

# reshape
hw <- hw %>%
  mutate(
    warm = case_when(
      warm == 1 ~ "Strongly disagree",
      warm == 2 ~ "Disagree",
      warm == 3 ~ "Agree",
      warm == 4 ~ "Strongly agree"
    ),
    warm = factor(
      warm,
      levels = c(
        "Strongly disagree",
        "Disagree",
        "Agree",
        "Strongly agree"
      ),
      ordered = TRUE
    )
  )

# check levels in data
levels(hw$warm)
```

**Question: Why would a multinomial model throw away useful information about this dependent variable? Why would treating `warm` as a continuous variable also be questionable?**

### Ordered Logit

We will estimate an ordered logistic regression using `clm()` from the `ordinal` package.

```{r}
# fit an ordered logit
warm_model <- clm(
  warm ~ male + white + ed + prst + age,
  data = hw,
  link = "logit"
)

# inspect results
summary(warm_model)
```

An ordered logit estimates both coefficients for the predictors and thresholds separating adjacent outcome categories. Under the proportional-odds formulation, each predictor has one coefficient that applies across all cumulative splits of the outcome.

Just like with logistic regression for binary outcomes, exponentiating a slope coefficient gives an odds ratio.

```{r}
# exponentiate coefficients
exp(coef(warm_model)[c("male", "white", "ed", "prst", "age")])
```

**Question: Choose one predictor and interpret its odds ratio.**

### Predicted Probabilities

With an ordered outcome, a single observation has a predicted probability for each response category. These probabilities should sum to 1. For this example, we will vary respondent age while holding the remaining covariates at a fixed profile.

```{r}
# generate prediction grid varying age
warm_grid <- tibble(
  age = seq(
    min(hw$age, na.rm = TRUE),
    max(hw$age, na.rm = TRUE),
    by = 1
  ),
  male = 0,
  white = 1,
  ed = mean(hw$ed, na.rm = TRUE),
  prst = mean(hw$prst, na.rm = TRUE)
)

# generate predicted probability of each response category
warm_pred <- predict(
  warm_model,
  newdata = warm_grid,
  type = "prob",
  se.fit = TRUE,
  interval = TRUE,
  level = 0.95
)

# extract predicted probabilities
warm_probs <- warm_pred$fit

# check predicted probabilities sum to 1
head(rowSums(warm_probs))

# combine prediction grid and predicted probabilities
warm_predictions <- bind_cols(
  warm_grid,
  as.data.frame(warm_probs)
)

```

We'll present the predicted probabilities in two different ways. First, we'll ploy all categories as a function of one predictor (age) along with confidence intervals. Then, we'll plot the $J-1$ cumulative probabilities across age.

```{r}
# combine predicted probabilities and confidence intervals
warm_predictions_ci <- bind_cols(
  warm_grid,
  
  as.data.frame(warm_pred$fit) %>%
    rename_with(~ paste0(.x, "_fit")),
  
  as.data.frame(warm_pred$lwr) %>%
    rename_with(~ paste0(.x, "_lwr")),
  
  as.data.frame(warm_pred$upr) %>%
    rename_with(~ paste0(.x, "_upr"))
) %>%
  
  # convert to long format
  pivot_longer(
    cols = -c(age, male, white, ed, prst),
    names_to = c("response", ".value"),
    names_pattern = "(.*)_(fit|lwr|upr)$"
  ) %>%
  
  # keep substantive ordering of the response categories
  mutate(
    response = factor(
      response,
      levels = c(
        "Strongly disagree",
        "Disagree",
        "Agree",
        "Strongly agree"
      )
    )
  )

# plot the predicted probabilities as a function of age
warm_predictions_ci %>%
  ggplot(
    aes(
      x = age,
      y = fit,
      color = response,
      fill = response
    )
  ) +
  geom_ribbon(
    aes(
      ymin = lwr,
      ymax = upr
    ),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 1) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(
    x = "Age",
    y = "Predicted Probability",
    color = "Response",
    fill = "Response"
  )

  # convert to long format
warm_predictions_long <- warm_predictions %>%
  pivot_longer(
    cols = -c(age, male, white, ed, prst),
    names_to = "response",
    values_to = "probability"
  ) %>%
  
  # keep substantive ordering of the response categories
  mutate(
    response = factor(
      response,
      levels = c(
        "Strongly disagree",
        "Disagree",
        "Agree",
        "Strongly agree"
      )
    )
  ) %>%
  
  # make upper and lower bounds
  arrange(age, response) %>%
  group_by(age) %>%
  mutate(
    ymax = cumsum(probability),
    ymin = lag(ymax, default = 0)
  ) %>%
  ungroup()

# plot predicted probabilities so that they collectively sum to 1
warm_predictions_long %>%
  ggplot(
    aes(
      x = age,
      ymin = ymin,
      ymax = ymax,
      fill = response,
      group = response
    )
  ) +
  geom_ribbon(
    linewidth = 0,
    color = NA
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_x_continuous(
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_viridis_d() +
  theme_bw() +
  labs(
    x = "Age",
    y = "Predicted Probability",
    fill = "Response"
  )
```

**Question: How does age change the distribution of predicted responses? Which categories become more likely as age increases, and which become less likely?**

### The Proportional-Odds Assumption

As we discussed in lecture, an ordered logit assumes that the relationship between a predictor and the outcome is the same across the cumulative splits of the dependent variable. One way to explore informally test this is to estimate separate binary logits at different thresholds.

For example, for `male`:

```{r}
warm_prop_1 <- glm(
  I(as.numeric(warm) >= 2) ~ male,
  family = binomial(link = "logit"),
  data = hw
)

warm_prop_2 <- glm(
  I(as.numeric(warm) >= 3) ~ male,
  family = binomial(link = "logit"),
  data = hw
)

warm_prop_3 <- glm(
  I(as.numeric(warm) >= 4) ~ male,
  family = binomial(link = "logit"),
  data = hw
)

coef(warm_prop_1)["male"]
coef(warm_prop_2)["male"]
coef(warm_prop_3)["male"]
```

**Question: How similar are these three coefficients? What would large differences suggest about the proportional-odds assumption?**

## Multinomial Outcomes

Lastly, we'll look at unordered categorical dependent variable with more than two categories, or a multinomial outcome. In this case, the values represent different alternatives but there is no meaningful ordering among them.

We will use data from the 2017/18 round of LAPOP in Brazil and examine reported vote choice among three candidates in the 2018 presidential election: Jair Bolsonaro, Fernando Haddad, and Ciro Gomes. There is no defensible ordering such as Bolsonaro < Haddad < Gomes, so an ordered model would not be appropriate.
The predictors are:

- `sex`: male (1) or female (0)
- `age`: respondent age
- `capital_pun`: support for capital punishment (1) or not (0)

Load the `BrazilLAPOP_2018.rds` data and clean it to fit a model. 

```{r}
# load data
brazil_data <- read_rds("BrazilLAPOP_2018.rds")

# filter data
brazil_data <- subset(brazil_data, 
                      vote_choice == "1501" | 
                        vote_choice == "1502" |
                        vote_choice == "1503")
brazil_data$vote_choice <- as.factor(brazil_data$vote_choice)

# rename levels to candidate names
brazil_data$vote_choice <- factor(
  brazil_data$vote_choice,
  levels = c("1501", "1502", "1503"),
  labels = c("Bolsonaro", "Haddad", "Gomes")
)


```

The first factor level is the reference category. With the ordering above, the model will estimate separate equations comparing Haddad with Bolsonaro and Gomes with Bolsonaro.

**Question: Why do we need a reference category in a multinomial logit model? What does it mean for interpreting the model coefficients?**

### Multinomial Logit

Fit a multinomial logit using `multinom()` from `nnet`.

```{r}
# fit a multinomial logit model
multi_model <- multinom(
  vote_choice 
  ~ sex 
  + age
  + capital_pun,
  data = brazil_data,
  Hess = TRUE,
  trace = FALSE
)

# inspect results
summary(multi_model)
```

The coefficient matrix now contains one set of coefficients for each non-reference outcome. Each coefficient describes a change in the log odds of that outcome *relative to the reference outcome*. 

*Note:* You can change the reference category to whichever outcome is most useful for interpretation by resetting the factor’s reference level before fitting the model, for example with `relevel()`.

```{r}
# exponentiate coefficients
exp(coef(multi_model))
```

Exponentiating the coefficients converts them to relative risk ratios. ** say more about these **

**Question: Look at the coefficient on `age` in the Haddad equation. What two electoral outcomes are being compared? Then look at `age` in the Gomes equation. Why should we be cautious about trying to summarize the effect of age using either coefficient alone?**

### Predicted Probabilities

As with ordered logit, predicted probabilities are usually much easier to interpret than the raw coefficients. We will calculate the probability of supporting each candidate across respondent age while holding `sex = 0` and `capital_pun = 0`.

```{r}
# create grid of hypothetical observations
brazil_grid <- tibble(
  age = seq(
    min(brazil_data$age, na.rm = TRUE),
    max(brazil_data$age, na.rm = TRUE),
    length.out = 100
  ),
  sex = 0,
  capital_pun = 0
)

# generate predicted probabilities
brazil_probs <- predict(
  multi_model,
  newdata = brazil_grid,
  type = "probs"
)

# Confirm that probabilities sum to 1
head(rowSums(brazil_probs))

# combine observations and predicted probabilities
brazil_predictions <- bind_cols(
  brazil_grid,
  as.data.frame(brazil_probs)
)
```

Reshape and plot the probabilities.

```{r}
# reshape
brazil_predictions_long <- brazil_predictions %>%
  pivot_longer(
    cols = -c(age, sex, capital_pun),
    names_to = "candidate",
    values_to = "probability"
  )

# plot
brazil_predictions_long %>%
  ggplot(aes(
    x = age,
    y = probability,
    linetype = candidate
  )) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    x = "Age",
    y = "Predicted Probability of Vote Choice",
    linetype = "Candidate"
  )
```

**Question: Describe the substantive relationship between age and candidate preference. How does this plot help you understand the age coefficients from the model output?**






